---
format: gfm
editor: source
---

# Fetch and Merge Accelerometer data to Locations

MoveApps

Github repository: <https://github.com/dmpstats/Fetch_and_Bind_Acceleration_to_Locations>

## Description

Downloads available accelerometer data for individuals and time window comprised
in the input location data and merges acceleration measurements to each observed
location based on recorded time.

## Documentation {#sec-documentation}

Given the individuals and time-span of recorded events in the input location
data, this App is tasked with downloading and merging available (non-location)
acceleration data to recorded locations. Merging is entirely time-based, and the
underlying data processing can be outlined as follows:

1. Extract animal/track IDs and start-end timestamps of location events for
individuals present in input data;

2. Check user's download permissions and availability of accelerometer (ACC) data
for each individual;

3. Download sought ACC data via `move2::download_study_data()`;

4. Process ACC data into a standardized format. ACC samples can be recorded in
Movebank in plain format (one ACC measurement on each active axis per row) or
nested format (burst of consecutive AAC values stored per row). In this step:

    - Downloaded data is reshaped to return bursts of ACC values nested in a
    list-column named `acc_bursts`. Therefore, each row of standardized ACC
    represents an event at which a burst of ACC values (per active axis) was
    recorded, with the timestamp column specifying the starting time of the burst.
  
    - The type (raw or non-raw) and tag origin (eobs or non-eobs) of ACC samples 
    are also added as data attributes. 
  
    - ACC data can optionally be thinned down to only retain a single ACC burst
    event within a given time interval via the parameter `acc_timefilter` (details
    in the Settings section below). This feature can be useful to avoid long
    computations, and potentially lack of memory issues, when processing
    high-frequency ACC data (e.g. 1 burst every couple of seconds).


5. Merge processed ACC events to location events of the same animal/track based
on recording timestamps and according to a choice between two merging criteria:

    - `'latest'`: ACC events are allocated to the latest location event preceding 
    the starting time at which ACC bursts were recorded
  
    - `'nearest'`: ACC events are allocated to the closest-in-time location event, 
    which can occur either before or after the ACC sampling starting time.
    
    ACC events are joined to corresponding location events in a list-column
    named `acc_dt`. Data nesting is required here as, depending on the frequency
    of ACC events relative to location events, consecutive ACC events may be
    allocated to the same location event.


6. Prepare merged data for output, where `output` is a `move2` location object
containing the entire input data along with the following additions:

    - *event data*: the list-column `acc_dt` comprising the downloaded and merged 
    ACC events
    
    - *track data*: columns `acc_dwn_start_time`, `acc_dwn_end_time` and 
    `acc_in_timespan` providing information about the ACC downloading step.
    
    - *object attributes*: 
        - `acc_merging_rule` specifying the selected merging rule[^1]
        - `acc_track_data` providing the track data component of the merged 
        ACC data
        
    [^1]: retrievable via `attr(output, "acc_merging_rule")`
    


<br />

::: {.callout-note}
The creation of this App was prompted by a current limitation in MoveApps, which
allows Workflows to be initiated from either location or non-location data but
not both simultaneously. MoveApps has plans to ease this restriction in the
future, such as enabling multiple data inputs on Apps. Consequently, it is
possible that this App may become obsolete in the future.
:::

<br />


### Input data

`move2` location object


### Output data

`move2` location object - see point 6. of [Documentation](#sec-documentation)
for additional context. Section [Using the Output data](#sec-demo) demonstrates
how to access and manipulate the merged data in the output object.


### Artefacts

`downloaded_acc_data.rds`: a `tibble` object with downloaded and processed ACC
data, with ACC events and track data nested by animal in, respectively,
list-columns `acc_dt` and `acc_trk_dt`.




### Parameters

**Movebank username** (`usr`): character string, the Movebank username. Default: `NULL`.

**Movebank password** (`pwd`): character string, the Movebank password. Default: `NULL`.

**Merging Rule** (`merging_rule`): specify how downloaded Accelerometer
data is merged to location data. ACC events can allocated to either:

  - the latest recorded location available preceding sampling time (`'latest'`, 
  the default), or
  - the closest-in-time recorded location (`'nearest'`)

**Store ACC track information** (`store_acc_track_info`): check-box to choose
whether to store track data from merged ACC data as an attribute of the output
`move2` object. Default: `FALSE`.

**Filter downloaded ACC data by time interval** (`acc_timefilter`): an integer
defining the time interval, in minutes, for thinning the ACC data. Must be
between 0 (no filtering) and 30. Unit: `minutes`; default: `0`



### Most common errors

None yet



### Null or error handling

**data**: the App will skip the merging step and return a modified version of the input
data with an empty column `acc_dt` if:

- Accelerometer data is not collected for any of the animals in the input data,
- the user does not have downloading permission for the study

An informative warning is also issued.


**Movebank username** (`usr`) and **Movebank password** (`pwd`): If one of the
credentials are either NULL or connection to Movebank is voided due to invalid
log-in details, the app will return an error reminding the user to provide valid
credentials.

**Filter downloaded ACC data by time interval** (`acc_timefilter`): The App will 
throw an error if the specified interval is not between 0 and 30.



### Using the output data {#sec-demo}

The purpose of this App is to retrieve and combine available non-location
Accelerometer (ACC) data with the provided input location data. The format of
the returned output data is structured with data nesting, providing a versatile
approach for handling the merged information in subsequent applications. The
trade-off for this flexibility is the slightly increased level of data
manipulation needed to work with nested datasets.

Here we suggest some sample code to inspect, manipulate and apply the output
data effectively in e.g. downstream Apps within a MoveApps Workflow. Let's start
by loading the dependencies required to run this demonstration.

```{r}
#| message: false
#| warning: false

# load dependencies
library(move2)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(stringr)
library(gganimate)

source("RFunction.R")
source(file.path("src/common/logger.R"))
source(file.path("src/io/app_files.R"))
source(file.path("src/io/io_handler.R"))
```


```{r}
#| echo: false
#| include: false

options(pillar.width = 500)

# movebank credentials stored in HOME .Renviron file
usr <- Sys.getenv("vult_usr")
pwd <- Sys.getenv("vult_pwd")

data <- readRDS("data/raw/gps_vult_sa.rds") |>
  filter_track_data(individual_local_identifier == "Bateleur_8887") |>
  slice(1:10)
```

Now we run the App's underpinning `rFunction` for a given input `data`
using the `'lates'` merging rule.

```{r}
output <- rFunction(data, usr = usr, pwd = pwd, merging_rule = 'latest') 
```

The output data is a `move2` location object, where the merged Accelerometer
(ACC) data is provided as `tibble` objects nested in list-column `acc_dt`.

```{r}
output <- output |> 
  dplyr::select(where(~all(!is.na(.)))) # dropping redundant columns

output
```

#### Accessing nested ACC data

ACC events recorded e.g. between the second and third location points can
retrieved by

```{r}
output |> 
  as_tibble() |> 
  slice(2) |> 
  dplyr::select(acc_dt) |> 
  unnest(acc_dt)
```


Acceleration measurements recorded in each ACC event are nested in list-column
`acc_burst`. Thus, the burst of ACC values in each active axis sampled in the
first ACC event between the second and third location points is accessible via

```{r}
output |> 
  as_tibble() |> 
  slice(2) |> 
  dplyr::select(acc_dt) |> 
  unnest(acc_dt) |> 
  slice(1) |> 
  mutate(acc_burst = map(acc_burst, as_tibble)) |> # required for nicer unnesting compared to the original matrix format
  dplyr::select(acc_burst)|> 
  unnest(acc_burst)
```


#### `'latest'` versus `'nearest'` merging

@fig-merging-rule provide a graphical representation of the difference between
the two available options for merging the ACC data to location data (via
parameter `merging_rule`).

```{r}
#| echo: false
#| include: false

dt <- readRDS("data/raw/gps_vult_gaia.rds") |>
  mt_filter_per_interval(unit = "3 secs") |> 
  filter_track_data(individual_local_identifier == "V051") |>
  slice(1:10)

# merging via the nearest rule
output_nearest <- rFunction(dt, usr = usr, pwd = pwd, merging_rule = 'nearest') |>
  dplyr::select(where(~all(!is.na(.)))) |> # dropping redundant columns
  as_tibble() |> 
  mutate(merging = 'nearest')

output_latest <- rFunction(dt, usr = usr, pwd = pwd, merging_rule = 'latest') |>
  dplyr::select(where(~all(!is.na(.)))) |> # dropping redundant columns
  as_tibble() |> 
  mutate(merging = 'latest')


output_merg <- bind_rows(output_nearest, output_latest) |> 
  mutate(event_id = str_sub(event_id,-5,-1)) |> 
  group_by(merging)

```



```{r}
#| code-fold: true
#| out-width: 100%
#| fig-width: 10
#| fig-height: 7
#| label: fig-merging-rule
#| fig-cap: Comparing the `'latest'` and `'nearest'` rules for merging ACC data with location data. Vertical lines mark the timestamps of six location events, while points represent acceleration values sampled in ACC events. Points are colored according to their assigned location given the selected merging rule.

output_merg_unpacked <- output_merg |>
  as_tibble() |>
  mutate(acc_dt = map(acc_dt, \(acc_events){
    if(!is.null(acc_events)){
      acc_events |>
        mutate(
          acc_burst = pmap(
            list(eobs_acceleration_sampling_frequency_per_axis, timestamp, acc_burst),
            \(freq, start_time, acc){
              # downsize to only first 5 measurements on each acc event, for plotting clarity
              n <- 5
              acc |>
                as_tibble() |>
                slice(floor(seq(1, nrow(acc), length.out = n))) |>
                mutate(
                  acc_timestamp = start_time,
                  #acc_timestamp = seq.POSIXt(from = start_time, by = 1/freq, length.out = n),
                  .before = 1
                )
            })
        ) |>
        unnest(acc_burst) |>
        rename(acc_event_id = event_id) |>
        dplyr::select(acc_event_id, acc_timestamp, matches("acc_[xyz]"))
    } else NULL
  })) |>
  unnest(acc_dt)


location_times <- output_merg |> 
  ungroup() |> 
  distinct(event_id, geometry, timestamp) |> 
  mutate(timestamp = timestamp + 0.15) # tiny nudge to ease vizualization

p <- output_merg_unpacked |>
  pivot_longer(cols = c(acc_x, acc_y, acc_z), names_to = "acc_axis", values_to = "acc_value") |>
  ggplot() +
  geom_point(aes(x = acc_timestamp, y = acc_value, col = factor(event_id), group = merging), size = 2) +
  facet_wrap(~acc_axis, ncol = 1, scales = "free_y") +
  geom_vline(aes(xintercept = timestamp, col = factor(event_id)), data = location_times, linewidth = 1) +
  labs(title = "merging_rule: 'latest'", y = "acceleration", x = "timestamp",
       col = "Location event ID") +
  scale_color_brewer(type = "qual", palette = "Paired") +
  theme_light() +
  theme(legend.position = "bottom")


p + 
  transition_states(merging, transition_length = 1, state_length = 3) +
  ggtitle("Merging Rule: '{closest_state}'") +
  enter_fade() + 
  exit_shrink() +
  ease_aes('sine-in-out')

```




#### Summarising ACC measurements relative to location events

Summarizing acceleration values between consecutive location points may be
required in downstream applications such as animal behavior classifiers.

For instance, the variance of all acceleration values recorded
between each location event (per active ACC axis) can be computed as follows.

```{r}
loc_acc_var <- output |> 
  mutate(
    var = map(acc_dt, \(acc_events){
      if(!is.null(acc_events$acc_burst)){
        # bind list of matrices (one per ACC event) by row
        all_bursts <- do.call(rbind, acc_events$acc_burst)
        # variance in each active ACC axis
        apply(all_bursts, 2, var)  
      } else{
        NULL
      }
    })
  ) |> 
  unnest_wider(var, names_sep = "_")

loc_acc_var
```

Note that `loc_acc_var` is not a `move2` object anymore. This is a by-product of
the un-nesting process, which drops the `move2` and `sf` classes and their
attributes. However, back-conversion is straightforward using `{move2}`
functionality and the original `output` object, as e.g.

```{r}
loc_acc_var <- loc_acc_var |> 
  mt_as_move2(
    time_column = mt_time_column(output), 
    track_id_column = mt_track_id_column(output)
)

mt_is_move2(loc_acc_var)
```

